---
title: "BioOptical Phenotype Fiji 2025 Expedition"
author: "____"
date: "8/21/2025"
output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
---

# Setup

```{r Setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


This script organizes the BioOptical phenotype data of corals collected in the Fiji June 2025 expedition... add details.
test

### Load Packages
```{r, echo=F, include=F}
##Install Packages if Needed
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")

##Load Packages
library(ggplot2) #Required for ggplots
library(dplyr)
library(gplots)
library(gridExtra)
library(car)
library(vegan)
library(reshape)
library(reshape2)
library(broom)
library(pvclust)
library(dendextend)
library(colorspace)
library(circlize)
library(viridis)
library(caret)
library(tidyverse)
library(tidyr)
library(igraph)
library(fs)
#library(DescTools)


```

# Functions
```{r functions}

movingAverage <- function(x, n=1, centered=FALSE) 
{	if (centered) {
        before <- floor  ((n-1)/2)
        after  <- ceiling((n-1)/2)
    } else {
        before <- n-1
        after  <- 0
    }
    s     <- rep(0, length(x))
    count <- rep(0, length(x))
	new <- x
    count <- count + !is.na(new)
    new[is.na(new)] <- 0
    s <- s + new
	i <- 1
    while (i <= before) {
        new   <- c(rep(NA, i), x[1:(length(x)-i)])
		count <- count + !is.na(new)
        new[is.na(new)] <- 0
        s <- s + new
        i <- i+1
    }
    i <- 1
    while (i <= after) {
        new   <- c(x[(i+1):length(x)], rep(NA, i))
        count <- count + !is.na(new)
        new[is.na(new)] <- 0
        s <- s + new
        i <- i+1
    }
    s/count
}
```


# Loading Metadata 
```{r load metadata}
# Read the mapping file
sample_metadata <- read.csv( "data/ESA_coral_data_entry_2026 - Collections.csv", check.names = F)
sample_metadata$Site <- substr(sample_metadata$Site, nchar(sample_metadata$Site)-1, nchar(sample_metadata$Site))
sample_metadata$Site <- as.integer(sample_metadata$Site)
sample_metadata<- sample_metadata[,-15]

mapplist <- read.csv("data/MAPP_Metadata - Sheet1.csv", check.names = F)
colnames(sample_metadata)[13] <- 'Bag_number'
metadata <- sample_metadata %>% full_join(mapplist)

metadata <- metadata %>% dplyr::filter(!(is.na(Sophie)))
colnames(metadata)[21] <- 'CoralID'
metadata$CoralID <- tolower(metadata$CoralID)
write.csv(metadata, file = "Outputs/AS26_MAPPmetadata.csv", row.names = F)
sample_metadata <- read.csv( "Outputs/AS26_MAPPmetadata.csv")

sample_metadata <- sample_metadata %>%
  mutate(UniqueID = case_when(
    Date != '20260211'  & Project == 'ESA' ~ paste0('t', Site, "_", CoralID),
    Date != '20260211' & Project == 'OCC' ~ paste0('occ', Site, "_", CoralID),
    TRUE ~ CoralID
  ))
```
# Phenotyping
### Load and Compile Processed Phenotype Data
```{r compiling processed pheno}
###### compiling all photophys to one csv ################# 
# Set the directory containing your data files
#data_dir <- 'Data/alldatacomparison/All_Phenotyping_Data/Processed/'
data_dir <- 'data/all'

# Get a list of all files in the directory
file_list <- list.files(path = data_dir, full.names = TRUE)

# Initialize an empty list to store data frames
photophys <- list()
data <- read.csv(file_list[1])
photophys <- data[,2:6]
mapplist <- list()

# Loop through each file, read it, and store it in the list
for (file in file_list) {
  # Read the file 
  data <- read.csv(file)
  base_sample_id <- substr(path_file(file), 1, nchar(path_file(file)) - 18)
  mapp_no <- substr(path_file(file), nchar(path_file(file)) - 16, nchar(path_file(file)) - 14)
# CREATE A UNIQUE KEY (e.g., "Sample123_M01")
  unique_key <- paste0(base_sample_id, "_", mapp_no)

  # Store using the unique key
  photophys[[unique_key]] <- data[, 7]
  mapplist[[unique_key]] <- mapp_no
}

# Convert mapp_list to a dataframe as you did before
mapplist_df <- data.frame(
  SampleID = names(mapplist),
  mapp_no = as.numeric(unlist(mapplist)),
  # Use regex to pull the original CoralID back out if needed
  UniqueID = sub("_[^_]+$", "", names(mapplist)), 
  stringsAsFactors = FALSE
)
# mapplist_df <- mapplist_df %>% 
#   mutate( CoralID = sub("^[^_]+_([^_]+)", "\\1", SampleID))
sample_metadata <- sample_metadata %>% full_join(mapplist_df, by = 'UniqueID') 

write.csv(photophys, file = "Outputs/AS26_PhenotypingData.csv", row.names = TRUE)
photophys <- read.csv("Outputs/AS26_PhenotypingData.csv")
write.csv(sample_metadata, file = "Outputs/AS26_MAPPmetadata.csv", row.names = TRUE)
sample_metadata <- read.csv("Outputs/AS26_MAPPmetadata.csv", row.names = 1)
```
# Depth Gradient Phenotyping
### Load and Compile Processed Phenotype Data
```{r compiling processed depth}
###### compiling all photophys to one csv ################# 
# Set the directory containing your data files
#data_dir <- 'Data/alldatacomparison/All_Phenotyping_Data/Processed/'
data_dir <- 'data/depthgradient/Phenotyping/shallow/Processed'
mapper <- read.csv('data/depthgradient/02242026Collections.csv')
mapper$UniqueID <- tolower(mapper$Label)
mapper$CoralID <- paste0(mapper$Species, mapper$Genotype)
# Get a list of all files in the directory
file_list <- list.files(path = data_dir, full.names = TRUE)

# Initialize an empty list to store data frames
photophys <- list()
data <- read.csv(file_list[1])
photophys <- data[,2:6]
mapplist <- list()

# Loop through each file, read it, and store it in the list
for (file in file_list) {
  # Read the file 
  data <- read.csv(file)
  base_sample_id <- substr(path_file(file), 9, nchar(path_file(file)) - 18)
  mapp_no <- substr(path_file(file), nchar(path_file(file)) - 16, nchar(path_file(file)) - 14)
  base_sampler <- mapper$CoralID[mapper$UniqueID == base_sample_id]
# CREATE A UNIQUE KEY (e.g., "Sample123_M01")
  unique_key <- paste0(base_sampler, "_shallow")

  # Store using the unique key
  photophys[[unique_key]] <- data[, 7]
  mapplist[[unique_key]] <- mapp_no
}

# Convert mapp_list to a dataframe as you did before
mapplist_df <- data.frame(
  SampleID = names(mapplist),
  mapp_no = as.numeric(unlist(mapplist)),
  # Use regex to pull the original CoralID back out if needed
  CoralID = sub("_[^_]+$", "", names(mapplist)), 
  stringsAsFactors = FALSE
)
# mapplist_df <- mapplist_df %>% 
#   mutate( CoralID = sub("^[^_]+_([^_]+)", "\\1", SampleID))
mapper <- mapper %>% full_join(mapplist_df) 
shallowset <- photophys 
mappershallow <- mapper

photophys <- full_join(shallowset, deepset)
mapper <- full_join(mappershallow, mapperdeep)
mapper <- mapper %>% mutate(
  Date = case_when( 
  grepl('shallow', SampleID) ~ '20260224', 
  grepl('deep', SampleID) ~ '20260218'
   
  )
) 
write.csv(photophys, file = "Outputs/AS26_DepthGradientPhenotypingData.csv", row.names = TRUE)
photophys <- read.csv("Outputs/AS26_DepthGradientPhenotypingData.csv")
write.csv(mapper, file = "Outputs/AS26_DepthGradientmetadata.csv", row.names = TRUE)
depth_metadata <- read.csv("Outputs/AS26_DepthGradientmetadata.csv", row.names = 1)
colnames(depth_metadata)[1] <- 'Bag_number'
sample_metadata <- sample_metadata %>% mutate(
  CoralID = case_when(Project == 'DepthGradient' ~ paste0(Species, Genotype), 
                                                          TRUE ~ CoralID ) 
                      ) %>% dplyr::filter(Project == "DepthGradient")
sample_metadata1 <- sample_metadata %>% full_join(depth_metadata, by = c('Date', 'Species', 'Genotype', 'UniqueID', 'CoralID', 'Bag_number'))

```



# Reflectance 
```{r abs new kenny method}
process_abs <- function(file) { 
stream<- read.csv(file, header = F)
wavy <- data.frame(stream[-1, ])
wavy[,1] <- as.numeric(wavy[,1])
wavy <- wavy[!is.na(wavy[,1]),]

wave <- rownames(wavy)[as.numeric(wavy[,1]) >= 400 & as.numeric(wavy[,1]) <= 750] #wavelengths 400 -> 800 (dependent on spectrometer)

PARrow<-c(as.numeric(wave[1]), as.numeric(wave[length(wave)]))

#column names
ical<-list()
namerChk<-list()
for(I in 2:ncol(wavy)) {
  qualityChk<-mean(as.numeric(wavy[-1,I])[PARrow[2]:(PARrow[2]+10)])-100
  if(abs(qualityChk)<5) {
    RefDat<-log(abs(1/(movingAverage(as.numeric(wavy[-1,I]), 10, TRUE)/100))) ##Absorbance Conversion
    namerChk[[I]]<-stream[1,I]
    RefDatAdj<-(RefDat-mean(RefDat[PARrow[2]:(PARrow[2]+10)]))[PARrow[1]:PARrow[2]]
    ical[[I]]<-RefDatAdj
  }
}
fixedDat<-data.frame(do.call(cbind, ical))
namelist<-t(data.frame(do.call(cbind, namerChk)))
namelist <- sub("^0", "", namelist)
dater <- substr(basename(file), 1, 8)
if (dater %in% datelist) {
    if (file == "Data/Reflectance/06232025_RB_MW.csv") {
      namelist2 <- namelist
    } else {
      # Use the filtered namelist, not the full list
      cbass_F <- cbass %>% filter(date_string == dater)
      namelist2 <- cbass_F[match(namelist, cbass_F$Plug_ID), 5]
    }
  } else {
    namelist2 <- namelist
  }
######################
averDat<-t(aggregate(t(fixedDat), by=list(as.factor(namelist2)), FUN=mean, rm.na=TRUE))
colnames(averDat)<-averDat[1,]
averDat<-data.frame(averDat[-1,], check.names=F)
averDat<-data.frame("wave"=stream[PARrow[1]:PARrow[2],1], averDat)
rownames(averDat) <- 1:nrow(averDat)
######################
roir1<-rownames(averDat)[as.numeric(averDat[,1]) >= 580]
roir11<-c(as.numeric(roir1[1]), as.numeric(roir1[length(roir1)]))
roir2<-rownames(averDat)[as.numeric(averDat[,1]) >= 580 & as.numeric(averDat[,1]) <= 650]
roir22<-c(as.numeric(roir2[1]), as.numeric(roir2[length(roir2)]))



fin<-list()
for(D in 2:dim(averDat)[2])
{
  chlaMax<-averDat[roir11[1]:roir11[2],1][which.max(averDat[roir11[1]:roir11[2],D])]
  chlaMin<-averDat[roir22[1]:roir22[2],1][which.min(averDat[roir22[1]:roir22[2],D])]
  cMax<-as.numeric(rownames(averDat[averDat$wave == chlaMax, ]))
  cMin<-as.numeric(rownames(averDat[averDat$wave == chlaMin, ]))
  ChloroPeak<-mean(as.numeric(averDat[(cMax-6):(cMax+6),D]))
  ChloroAUC<-AUC(x=as.numeric(averDat[(cMax-12):(cMax+12),1]), y=as.numeric(averDat[(cMax-12):(cMax+12),D]), method="trapezoid")
  ChlcaAUC<-AUC(x=as.numeric(averDat[cMin:dim(averDat)[1],1]), y=as.numeric(averDat[cMin:dim(averDat)[1],D]), method="trapezoid")
  WholeAUC<-AUC(x=as.numeric(averDat[,1]), y=as.numeric(averDat[,D]), method="trapezoid")
  CBASS_ID<-colnames(averDat)[D]
  fin[[D]]<-data.frame(CBASS_ID, ChloroPeak, ChloroAUC, ChlcaAUC, WholeAUC, "cMaxnm"=averDat[cMax,1])
}
finished<-data.frame(do.call(rbind, fin))   #this contains all the data processed and ready to be used to calculate % change between conrtol and treatment, or across time points.
return(finished)
}

# -----------------------------------------
# Example: process all CSVs in a folder
# -----------------------------------------
folder <- "Data/Reflectance"   # change to your folder
files <- list.files(folder, pattern = "\\.csv$", full.names = TRUE)
abf <- 0
for(file in files) { 
a <- process_abs(file)
if(abf == 0) { 
  reflectance <- a 
  abf = 1 
  } else { 
  reflectance <- rbind(reflectance, a) } 
}
#colnames(reflectance) <- tolower(colnames(reflectance))

col_names <- reflectance$CBASS_ID
# Find the names that are duplicated
duplicates <- col_names[duplicated(col_names)]
# Print the unique duplicated names
duplicate_indexes <- which(col_names %in% duplicates)
reflectance <- reflectance[-duplicate_indexes,]
reflectance$CBASS_ID <-trimws(reflectance$CBASS_ID)
reflectance$CBASS_ID <- tolower(reflectance$CBASS_ID)
write_csv(reflectance, file = "Outputs/Reflectance_summarized.csv")

```

### Percent Change Absorbance
Calculated by taking the percent change in absorbance between the control and heat treated fragments post-CBASS bleaching. 
```{r abs pchange}
## ONLY INCLUDES CBASS ###### 
reflectance <- read.csv('Outputs/Reflectance_summarized.csv', check.names = F)
column_names <- colnames(reflectance)
cbass$CoralID <- tolower(cbass$CoralID)
factorlevels <- levels(as.factor(cbass$CoralID))
perchange<-list()
coldspectra <- list()
hotspectra <- list()

for (colony in factorlevels) {
  hot<-paste0(colony, "h")
  cold<-paste0(colony, "c")
  
  hotspectra <- reflectance[reflectance$CBASS_ID == hot,]
  coldspectra <- reflectance[reflectance$CBASS_ID == cold,]
  
  if (nrow(hotspectra)!=0 && nrow(coldspectra)!=0) {
    perchange[[colony]] <- data.frame( 
       pC_ChloroPeak = ((hotspectra$ChloroPeak - coldspectra$ChloroPeak)/coldspectra$ChloroPeak) * 100, 
      pC_ChloroAUC = ((hotspectra$ChloroAUC - coldspectra$ChloroAUC)/coldspectra$ChloroAUC) * 100, 
      pC_ChlcaAUC = ((hotspectra$ChlcaAUC - coldspectra$ChlcaAUC)/coldspectra$ChlcaAUC) * 100,
      pC_WholeAUC = ((hotspectra$WholeAUC - coldspectra$WholeAUC)/coldspectra$WholeAUC) * 100      
      )
  }
}
allDater <-data.frame(do.call(rbind, perchange)) 
dim(allDater)
refl_pC <- allDater
refl_pC$CoralID <- rownames(refl_pC)

## missing IDS 
missing <- cbass[!(cbass$CoralID %in% refl_pC$CoralID),]
print("CoralID's missing from final reflectance dataset")
unique(missing$CoralID)


write.csv(refl_pC, file = 'Outputs/Reflectance_PC.csv', row.names = FALSE)
refl_pC <- read.csv('Outputs/Reflectance_PC.csv')


```

# Reflectance Old Method 

### Load and Organize Reflectance Data
```{r}
#### load files ####

#### calculate absorbance #### 
process_abs_old <- function(file) { 
  name <- read.csv(file, header = F)
  #name <- day10
  wave<-as.numeric(name[-1,1]) 
  valid_idx <- which(wave > 400 & wave < 800)
  wave2 <- wave[valid_idx]
  namer<-t(name[1,-1]) #column names

  ical<-list() 
  namerChk<-list() 
  
  for(I in 2:ncol(name)) {	
    data_col <- as.numeric(name[-1, I])[valid_idx]
    qualityChk <- mean(tail(data_col, 10)) - 100
    
    if(abs(qualityChk)<5) {	
      RefDat <- log(abs(1 / (movingAverage(data_col, 10, TRUE) / 100)))
      namerChk[[I]] <- name[1, I]
      RefDatAdj <- (RefDat - mean(tail(RefDat, 10)))
      ical[[I]] <- RefDatAdj
    }
  }
  fixedDat<-data.frame(do.call(cbind, ical))
  namelist<-t(data.frame(do.call(cbind, namerChk)))
  namelist <- sub("^0", "", namelist)
  dater <- substr(basename(file), 1, 8)
  
if (dater %in% datelist) {
    if (file == "Data/Reflectance/06232025_RB_MW.csv") {
      namelist2 <- namelist
    } else {
      # Use the filtered namelist, not the full list
      cbass_F <- cbass %>% filter(date_string == dater)
      namelist2 <- cbass_F[match(namelist, cbass_F$Plug_ID), 5]
    }
  } else {
    namelist2 <- namelist
  }
  ##################
  averDat<-t(aggregate(t(fixedDat), by=list(as.factor(namelist2)), FUN=mean, rm.na=TRUE))
  colnames(averDat)<-averDat[1,]
  ##################
  controlT3a<-data.frame(averDat[-1,], check.names = F)
  rownames(controlT3a)<-wave2
  t3a_num <- as.data.frame(lapply(controlT3a, as.numeric))
  colnames(t3a_num)<- colnames(controlT3a)
  controlT3a <- t3a_num
  controlT3a$wave <- wave2
  controlT3a$wave <- round(controlT3a$wave, digits = 0) 
  colnames(controlT3a) <-trimws(colnames(controlT3a))
  colnames(controlT3a) <- tolower(colnames(controlT3a))
  controlT3a <- controlT3a %>% 
    group_by(wave) %>% 
    summarise(across(where(is.numeric), mean, na.rm = TRUE ))
  return(controlT3a)
}

#### final reflectance dataframe #### 
folder <- "Data/Reflectance"   # change to your folder
files <- list.files(folder, pattern = "\\.csv$", full.names = TRUE)
abf <- 0
for(file in files) { 
a <- process_abs_old(file)
print(file)
if(abf == 0) { 
  reflectance <- a 
  abf = 1 
  } else { 
  reflectance <- reflectance %>% full_join(a, by = 'wave') } 
}
backup <- reflectance
reflectance <- backup
####### averaging duplicates ########### 
clean_and_average_duplicates <- function(df) {
  
  # 1. Get all column names
  cols <- colnames(df)
  
  # 2. Identify unique base names (e.g., 'SampleA' from 'SampleA.x' or 'SampleA.1')
  # This regex handles .x, .y, and .[digit] suffixes
  base_names <- str_replace(cols, "\\.[xy1]$|\\.\\d+$", "")
  
  # Find base names that appear more than once (the duplicates)
  duplicate_base_names <- base_names[duplicated(base_names) | duplicated(base_names, fromLast = TRUE)] %>%
    unique()
  
  df_clean <- df
  
  # 3. Loop through each duplicated base name and calculate the average
  for (base_name in duplicate_base_names) {
    
    # Find all columns that match this base name, including suffixes
    matching_cols <- cols[str_detect(cols, paste0("^", base_name, "(\\.[xy1]$|\\.\\d+$)")) | cols == base_name]
    
    # Filter only the suffixed columns for averaging (e.g., SampleA.x, SampleA.y)
    suffixed_to_average <- matching_cols[str_detect(matching_cols, "\\.[xy1]$|\\.\\d+$")]
    
    if (length(suffixed_to_average) > 0) {
      
      # Use rowwise() and c_across() to calculate the mean across the row
      df_clean <- df_clean %>%
        rowwise() %>% 
        mutate(
          # Create the new averaged column with the base name
          !!base_name := mean(c_across(all_of(suffixed_to_average)), na.rm = TRUE)
        ) %>%
        ungroup() %>%
        # Remove the original suffixed columns
        select(-all_of(suffixed_to_average))
    }
  }
  
  return(df_clean)
}

# Apply the function to your dataframe
combined_data_clean <- clean_and_average_duplicates(reflectance)




write.csv(combined_data_clean, file = "Outputs/F25_AbsorbanceData_oldmethod.csv", row.names = F)
```

# Percent Change Absorbance Old Method 
DO NOT USE IF USING NEWER ABSORBANCE CALCULATIONS
```{r}
reflectance <- read.csv('Outputs/F25_AbsorbanceData_oldmethod.csv')

######### CBASS ############## 
column_names <- colnames(reflectance)
factorlevels <- levels(as.factor(tolower(cbass$CoralID)))
perchange<-list()
coldspectra <- list()
hotspectra <- list()
reflectance_pC <- list()

for (colony in factorlevels) {
  hot<-paste0(colony, "h")
  cold<-paste0(colony, "c")

  hotspectra <- reflectance[[hot]]
  coldspectra <- reflectance[[cold]]
  
  if (!is.null(hotspectra) && !is.null(coldspectra)) {
    perchange <- ((hotspectra - coldspectra) / coldspectra) * (100)
    dataSpectra <- data.frame(reflectance$wave, coldspectra, hotspectra, perchange)
    reflectance_pC[[colony]] <- data.frame(
      "colony" = colony, 
      "nm415" = round(dataSpectra[(which(dataSpectra$reflectance.wave == 415)), 4]), 
      "nm448" = round(dataSpectra[(which(dataSpectra$reflectance.wave == 448)), 4]), 
      "nm470" = round(dataSpectra[(which(dataSpectra$reflectance.wave == 470)), 4]), 
      "nm505" = round(dataSpectra[(which(dataSpectra$reflectance.wave == 505)), 4]), 
      "nm675" = round(dataSpectra[(which(dataSpectra$reflectance.wave == 675)), 4])
    )
  }
}

allDater <-data.frame(do.call(rbind, reflectance_pC)) 
dim(allDater)
refl_pC <- allDater
colnames(refl_pC) <- c('CoralID', "nm415", "nm448", "nm470", "nm505", "nm675")

write.csv(refl_pC, file = 'Outputs/F25_Absorbance_CBASS_oldmethod_pC.csv', row.names = FALSE)
refl_pC <- read.csv('Outputs/F25_Absorbance_CBASS_oldmethod_pC.csv')

```

# FvFm data 

### Load and Compile Processed Endpoint Data
```{r compiling processed endpont}
###### compiling all fvfm to one csv ################# 
# Set the directory containing your data files
data_dir <- '/Users/sophiewong/Documents/GraduateSchool/Lab Notes/Fiji2025/Data/FvFm'

# Get a list of all files in the directory
file_list <- list.files(path = data_dir, full.names = TRUE)

# Initialize an empty list to store data frames
endpoint <- list()
data <- read.csv(file_list[1])
endpoint <- data[,2:6]

# Loop through each file, read it, and store it in the list
for (file in file_list) {
  # Read the file 
  data <- read.csv(file)
  base_sample_id <- substr(path_file(file), 1, nchar(path_file(file)) - 16)
 
  # Append the data frame to the list
  endpoint[[base_sample_id]] <- data[,7]
 
}
colnames(endpoint) <- tolower(colnames(endpoint))
write.csv(endpoint, file = "Outputs/F25_FvFmData.csv", row.names = TRUE)
endpoint <- read.csv("Outputs/F25_FvFmData.csv")


```
### calculating percent change FvFm 
```{r pchange FvFm}
endpoint <- read.csv('Outputs/F25_FvFmData.csv')
endpoint <- endpoint[,-c(1, 3:6)]
colnames(endpoint) <- tolower(colnames(endpoint))
 #colnames(endpoint)[1] <- 'CBASS_ID'

factorlevels <- levels(as.factor(cbass$CoralID))
endpoint_pC<-data.frame(endpoint$uni)
coldend <- list()
hotend <- list()

for (colony in factorlevels) {
  hot<-paste0(colony, "h")
  cold<-paste0(colony, "c")
  
  hotend <- endpoint[[hot]]
  coldend <- endpoint[[cold]]
  
  if (!is.null(hotend) && !is.null(coldend)) {
    perchange <- ((coldend - hotend) / coldend) * (-100)
    dataSpectra <- data.frame(coldend, hotend, perchange)
    endpoint_pC[[colony]] <- dataSpectra$perchange
  }
}

## missing IDS 
missing <- cbass[!(cbass$CoralID %in% colnames(endpoint_pC)),]
print("CoralID's missing from final reflectance dataset")
unique(missing$CoralID)



write.csv(endpoint_pC, file = 'Outputs/F25_endpoint_PC.csv')

endpoint_bycol <- endpoint_pC %>% 
  pivot_longer(
    cols = 2:ncol(endpoint_pC),
    names_to = "CoralID",
    values_to= "MetricValue"
  )
# Step 2: Reshape from long to wide, using endpoint.uni to create new columns
final_df <- endpoint_bycol %>%
  pivot_wider(
    names_from = endpoint.uni,
    values_from = MetricValue
  )
endpoint_bycol <- final_df %>% 
  select(CoralID, Quant.1.0, Quant.1.1, Quant.1.2, Quant.1.3)
colnames(endpoint_bycol) <- c("CoralID", "FvFm_415", "FvFm_448", "FvFm_470", "FvFm_505")
write.csv(endpoint_bycol, file = "Outputs/F25_FvFm_PC.csv")

```

Combining absorbance and fvfm dataframes for Ray 
```{r}
fvfm <- read.csv('Outputs/F25_FvFm_PC.csv', row.names = 1)
refl <- read.csv('Outputs/F25_Absorbance_CBASS_oldmethod_pC.csv')

endpoint <- fvfm %>% full_join(refl)
write.csv(endpoint, file = 'Outputs/F25_FvFm_Abs_oldmethod_pC.csv', row.names = F)
```

