# Setup

```{r Setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


This script creates a dendrogram of the bio-optical phenotype data from the compiled file created with ShinyProcessedFile_Compiler.R. 

### Load Packages
```{r, echo=FALSE, include=FALSE}

##Load Packages
library(ggplot2) #Required for ggplots
library(dplyr)
library(gplots)
library(gridExtra)
library(car)
library(vegan)
library(reshape)
library(reshape2)
library(broom)
library(pvclust)
library(dendextend)
library(colorspace)
library(circlize)
library(viridis)
library(caret)
library(tidyverse)
library(tidyr)
library(igraph)
library(fs)
library(alluvial)
library(ggalluvial)
library(umap)     
library(caret)
options(scipen=999)
set.seed(123)

```

# Loading Data 
```{r loading data, echo=F}
photophys <- read.csv("./Data/AS26_PhenotypingData_expt.csv", row.names = 1)

```

#OLD UMAP CODE
# Trajectory UMAPs 
## Photophys UMAP  

```{r photophys umap setup }
######### set up photophys dataframe ######## 
longphoto <- as.data.frame(t(photophys))
colnames(longphoto) <- longphoto[1,] 
longphoto <- as.data.frame(longphoto[-1:-5,]) # removing metrix, stage, col 
longphoto[,2:955] <- lapply(longphoto[,2:955], as.numeric)
longphoto$SampleID <- rownames(longphoto)
longphoto <- longphoto %>% relocate(SampleID)
fullyears <- longphoto 
fullyears <- data.frame(fullyears[, colSums(is.na(fullyears)) == 0])

allphoto <- fullyears %>% dplyr::select(-contains('Fo'), -contains('Fm'))
rownames(allphoto) <- allphoto$SampleID
photophys_clean <- allphoto %>% 
  dplyr::select(-c(SampleID)) # remove non-numeric columns 
photophys_clean <- photophys_clean %>% mutate(across(everything(), as.numeric))
allphoto[,2:ncol(allphoto)] <- allphoto[,2:ncol(allphoto)] %>% mutate(across(everything(), as.numeric))

photo.transit <- preProcess(x = photophys_clean, rangeBounds = c(0, 1), method = "range") # need to scale data 
photo_proc <- predict(photo.transit, photophys_clean)
photo_proc <- photo_proc[,-nearZeroVar(photo_proc)] # remove any columns with low variance across samples 
photo_proc <- photo_proc %>%
  dplyr::select(where(~ sum(is.na(.)) == 0))

photo_umapresult<-umap(photo_proc, n_neighbors= 15, min_dist= 0.1, metric="euclidean", n_components=2) # these parameters can change quite a bit-- worth looking up to figure out what to set them at for your data specifically 
plot(as.numeric(photo_umapresult$layout[,1]), as.numeric(photo_umapresult$layout[,2]), pch=19, cex=0.9, axes=T)


##### linking metadata to umap points ###### 


# --- Prep the Data for Lines and Points ---
umap_plotting_data <- data.frame(
  UMAP1 = photo_umapresult$layout[,1], 
  UMAP2 = photo_umapresult$layout[,2], 
  SampleID = allphoto$SampleID,
  Site = as.numeric(str_extract(allphoto$SampleID, "(?<=t)\\d+(?=_)")), 
  Colony =(str_extract(allphoto$SampleID, "(?<=_).+?(?=\\.)")))
umap_plotting_data=umap_plotting_data %>% 
  mutate(Species =as.factor(str_to_upper(substr(Colony,1,4))),
         Genotype=paste0(str_to_upper(Species),
                         "_s",formatC(Site,flag=0,width=2),
                         "_c",formatC(substr(Colony,5,99),flag=0,width=2)))

# --- Plotting Code: Spp specific plots ---
ggplot(umap_plotting_data, aes(x = UMAP1, y = UMAP2)) +
  # 2. Plot the points (Start and End Locations)
  geom_text(aes(label=Site, color=Species),
            size = 3,
            alpha = 0.8) +
  
  labs(title = "Photophysiology UMAP",
       x = "UMAP Dimension 1",
       y = "UMAP Dimension 2") +
  theme_minimal()
#write.csv(umap_results$layout, "Final_UMAP_Coords.csv")
#ggsave(filename = 'Figures/UMAP/trajectory_photophys_splitbyspp_coloredbygeno_chull.png', plot = last_plot(), width = 7.5, height = 7.5, units = "in")


```

# NEW UMAP CODE 
### Load Data and scale 
```{r}
set.seed(123)
longphoto <- as.data.frame(t(photophys))
colnames(longphoto) <- longphoto[1,] 
longphoto <- as.data.frame(longphoto[-1:-5,]) # removing metrix, stage, col 
longphoto[,2:955] <- lapply(longphoto[,2:955], as.numeric)
longphoto$SampleID <- rownames(longphoto)
longphoto <- longphoto %>% relocate(SampleID)
fullyears <- longphoto 
fullyears <- data.frame(fullyears[, colSums(is.na(fullyears)) == 0])

allphoto <- fullyears %>% dplyr::select(-contains('Fo'), -contains('Fm'))

##### dataset needs to have metrics in columns, samples in rows 
rownames(allphoto) <- allphoto$SampleID
photophys_clean <- allphoto %>% 
  dplyr::select(-c(SampleID)) # remove non-numeric columns

head(photophys_clean) 
names(photophys_clean)[colSums(is.na(photophys_clean)) > 0]
photophys_clean <- photophys_clean %>% mutate(across(everything(), as.numeric))
allphoto[,2:ncol(allphoto)] <- allphoto[,2:ncol(allphoto)] %>% mutate(across(everything(), as.numeric))

photophys_scaled <- photophys_clean %>%
  scale() %>% # scaling with zscores instead 
  as.data.frame()

#Sort out NAs in photophys_scaled data
dim(photophys_scaled)
NA_ij=which(is.na(photophys_scaled),arr.ind = T)
NA_ic=sort(table(NA_ij[,1]),decreasing = T);NA_ij#Every row has 102 NAs
NA_jc=sort(table(NA_ij[,2]),decreasing = T);NA_jc#There are 102 metrics that are full NA
NA_j=as.numeric(names(NA_jc))
NA_ppParams=names(photophys_scaled)[NA_jc]

#drop columns that end up NA in scaled data
photophys_scaled.=photophys_scaled %>% dplyr::select(-all_of(NA_ppParams))
dim(photophys_scaled.)
photophys_clean.=photophys_clean %>% dplyr::select(-all_of(NA_ppParams))
dim(photophys_scaled.)
allphoto.=allphoto %>% dplyr::select(-all_of(NA_ppParams))
dim(photophys_scaled.)
ppParam=names(photophys_scaled.)

```

#Incorporate ED50, and ID correlates
```{r}
ed=read.csv(file = "./Data/AS26_ED50_CI_BIC_TAO.csv")
allphoto.ed=allphoto. %>% mutate(
  Site = as.numeric(str_extract(SampleID, "(?<=t)\\d+(?=_)")), 
  Colony =(str_extract(SampleID, "(?<=_).+?(?=\\.)")),
  Species =as.factor(str_to_upper(substr(Colony,1,4))),
  Genotype=paste0(str_to_upper(Species),
                  "_s",formatC(Site,flag=0,width=2),
                  "_c",formatC(as.numeric(substr(Colony,5,6)),flag=0,width=2))) %>% 
  left_join(ed,by="Genotype") %>% 
  filter(ED50<40,ED50ci95<=2)
#allphoto.ed$ED50

#colinearity test
# 1. Calculate the full correlation matrix (highly optimized)
# Use "pairwise.complete.obs" if you have NAs in your sensor data
COLLINEAR_THRESHOLD=0.6
cor_matrix <- cor(allphoto.ed[, ppParam], use = "pairwise.complete.obs")

# 2. Flatten the matrix into a long-format table
# We use lower.tri to avoid duplicate pairs (A-B and B-A) and self-correlation (A-A)
cor_tri <- which(lower.tri(cor_matrix), arr.ind = TRUE)

cor_df <- data.frame(
  Sensor1 = rownames(cor_matrix)[cor_tri[, 1]],
  Sensor2 = colnames(cor_matrix)[cor_tri[, 2]],
  Correlation = cor_matrix[cor_tri]
)

#to_remove <- caret::findCorrelation(cor_matrix, cutoff = 0.8, exact = TRUE)
library(igraph)
diag(cor_matrix) <- 0
adj_matrix <- abs(cor_matrix) > COLLINEAR_THRESHOLD
adj_matrix[is.na(adj_matrix)] <- FALSE # Handle any remaining NAs

# 3. Create a graph where nodes are sensors and edges are "high correlation"
g <- graph_from_adjacency_matrix(adj_matrix, mode = "upper",diag=FALSE)

# 4. Instant Thinning: Pick one representative per cluster
# This is the fastest way to handle 3000+ variables
clusters <- components(g)$membership
ppParam_thinned <- ppParam[!duplicated(clusters)]

cat("Reduced from", length(ppParam), "to", length(ppParam_thinned), "sensors.")
# 4. Report results
cat("Original count:", length(ppParam), "\n")
cat("Thinned count: ", length(ppParam_thinned), "\n")

#Stability Selection with LASSO 
#Meinshausen, N., & Bühlmann, P. (2010). Stability selection. Journal of the Royal Statistical Society Series B: Statistical Methodology, 72(4), 417-473.
################################################
################################################
#Claude's Version: Sonnet 3.6
library(stabs)
library(glmnet)

# Prepare model matrix (no intercept, glmnet handles it)
X <- as.matrix(allphoto.ed[, ppParam_thinned])
y <- allphoto.ed[["ED50"]]

# Stability selection with LASSO base learner
#WITHOUT COLINEARITY CONTROL ppParam #3907
#cutoff = .75 and PFER = 1 returned no selected params
#cutoff = .6 and PFER = 5 returned no selected params
#WITH COLINEARITY CONTROL ppParam_thinned #184 cor threshold 0.7
#cutoff = .6 and PFER = 5 stably returned 4 selected param:  "Connect.20.2"       "Connect.19.3"       "range.4.rSigma.L2"  "L12CompL32.qTau1.4"
#cutoff = .75 and PFER = 1 returned no selected params


stab_result <- stabsel(
  x = X,
  y = y,
  fitfun = glmnet.lasso,
  cutoff = 0.6,      # selection probability threshold
  PFER = 5,# per-family error rate bound
  B=1000
)

# View results
print(stab_result)
plot(stab_result)

# Extract selected variables
selected_vars <- names(stab_result$selected)
cat("Selected variables:\n")
print(selected_vars)

#highlight other variables:
collinear_groups=setNames(vector("list", length(selected_vars)), selected_vars)
for(i in 1:length(selected_vars)){
  target=selected_vars[i]
  target_correlations <- cor_matrix[, target]
  
  # 2. Filter for those above your threshold (e.g., 0.6)
  # We exclude the target itself (where cor == 1)
  col_i=which(abs(target_correlations) > COLLINEAR_THRESHOLD)
  if(length(col_i)==0){collinear_group=NA}else{
    collinear_group=names(target_correlations)[col_i]
  }
  collinear_groups[[i]]=collinear_group
}
print(selected_vars)
print(collinear_groups)

ppParam_selected=selected_vars

ggplot(allphoto.ed,aes(x=.data[[ppParam_selected[1]]],
                    y=.data[[ppParam_selected[2]]],
                    shape=Species,color=ED50))+scale_color_viridis_c()+
  geom_point()
################################################

# ################################################
# #Tom's Janky LM run
# library(tidyverse)
# library(broom)
# ParamLM.DF=NULL
# bagfraction=.63 # same as full bootstrap. So why not bootstrap?
# Nrounds=5
# for (ri in 1:Nrounds){
#   allphoto.ed.bag=allphoto.ed[sample(1:nrow(allphoto.ed),size=floor(bagfraction*nrow(allphoto.ed))),]
#   ParamLM <- allphoto.ed.bag %>%
#     dplyr::select(all_of(c("ED50",ppParam))) %>% 
#     pivot_longer(cols = -ED50, names_to = "ppParam", values_to = "value") %>%
#     group_by(ppParam) %>%
#     do(tidy(lm(ED50 ~ value, data = .))) %>% 
#     filter(term=="value",!is.na(estimate))
#   
#   ParamLM$p.value.fdr=p.adjust(ParamLM$p.value,method = "fdr")
#   ParamLM$p.value.bon=p.adjust(ParamLM$p.value,method = "bonferroni")
#   ParamLM$Round=ri
#   
#   ParamLM.DF=rbind(ParamLM.DF,ParamLM)
#   print(ri)
# }
# alpha=0.05
# ppParam_4of5=ParamLM.DF %>% group_by(ppParam) %>% summarize(pSIG=sum(p.value<alpha)/length(p.value)) %>% 
#   filter(pSIG==.8) %>% pull(ppParam)
# 
# allphoto.ed %>% ggplot(aes(x=.data[[ppParam_4of5[sample(1:length(ppParam_4of5),1)]]],y=ED50))+geom_point()+stat_smooth(method="lm")
# allphoto.ed %>% ggplot(aes(x=.data[["L12CompR1.qqP.2"]],y=ED50))+geom_point()+stat_smooth(method="lm")
# ################################################

```

### evalulating parameters and clustering 
```{r}

# Test range of n_neighbors values
neighbor_vals <- c(5, 10, 15, 20, 30, 50)
min_dist_vals <- c(0.01, 0.05, 0.1, 0.2, 0.5)

photophys_scaled_thinned=photophys_scaled[,ppParam_thinned]
photophys_scaled_selected=photophys_scaled[,ppParam_selected]

umap_list <- lapply(neighbor_vals, function(n) {
  fit <- umap(photophys_scaled_thinned, n_neighbors = n, min_dist = 0.1,)
  as.data.frame(fit$layout) %>%
    setNames(c("UMAP1", "UMAP2")) %>%
    mutate(n_neighbors = n,
           SampleID= rownames(photophys_scaled_thinned),
           Site = as.numeric(str_extract(SampleID, "(?<=t)\\d+(?=_)")), 
           Colony =(str_extract(SampleID, "(?<=_).+?(?=\\.)")),
           Species =as.factor(str_to_upper(substr(Colony,1,4))),
           Genotype=paste0(str_to_upper(Species),
                           "_s",formatC(Site,flag=0,width=2),
                           "_c",formatC(as.numeric(substr(Colony,5,6)),flag=0,width=2)))
})

umap_list_4of5 <- lapply(neighbor_vals, function(n) {
  fit <- umap(photophys_scaled_4of5, n_neighbors = n, min_dist = 0.1,)
  as.data.frame(fit$layout) %>%
    setNames(c("UMAP1", "UMAP2")) %>%
    mutate(n_neighbors = n,
           SampleID= rownames(photophys_scaled),
           Site = as.numeric(str_extract(SampleID, "(?<=t)\\d+(?=_)")), 
           Colony =(str_extract(SampleID, "(?<=_).+?(?=\\.)")),
           Species =as.factor(str_to_upper(substr(Colony,1,4))),
           Genotype=paste0(str_to_upper(Species),
                           "_s",formatC(Site,flag=0,width=2),
                           "_c",formatC(as.numeric(substr(Colony,5,6)),flag=0,width=2)))
})

# Plot all versions for visual comparison
bind_rows(umap_list) %>%
  ggplot(aes(x = UMAP1, y = UMAP2, color = Species)) +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~n_neighbors, labeller = label_both, scales = "free") +
  theme_classic() +
  labs(title = "UMAP stability across n_neighbors values")

# Plot all versions for visual comparison
bind_rows(umap_list_4of5) %>%
  ggplot(aes(x = UMAP1, y = UMAP2, color = Species)) +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~n_neighbors, labeller = label_both, scales = "free") +
  theme_classic() +
  labs(title = "UMAP stability across n_neighbors values")

# Testing stability of parameter selection: Fit UMAP twice at each neighbor value with different seeds and compare cluster assignments
library(dbscan)
library(mclust)
stability_results <- lapply(neighbor_vals, function(n) {
  
  fit1 <- umap(photophys_scaled, n_neighbors = n, 
               min_dist = 0.1, random_state = 42)
  fit2 <- umap(photophys_scaled, n_neighbors = n, 
               min_dist = 0.1, random_state = 123)
  
  # Cluster both with DBSCAN -- objective clustering method, can change minPts and eps settings (will test eps settings later)
  db1 <- dbscan(fit1$layout, eps = 0.5, minPts = 5)$cluster
  db2 <- dbscan(fit2$layout, eps = 0.5, minPts = 5)$cluster
  
  data.frame(
    n_neighbors = n,
    ARI = adjustedRandIndex(db1, db2)
  )
}) %>%
  bind_rows()

stability_results_4of5 <- lapply(neighbor_vals, function(n) {
  
  fit1 <- umap(photophys_scaled_4of5, n_neighbors = n, 
               min_dist = 0.1, random_state = 42)
  fit2 <- umap(photophys_scaled_4of5, n_neighbors = n, 
               min_dist = 0.1, random_state = 123)
  
  # Cluster both with DBSCAN -- objective clustering method, can change minPts and eps settings (will test eps settings later)
  db1 <- dbscan(fit1$layout, eps = 0.5, minPts = 5)$cluster
  db2 <- dbscan(fit2$layout, eps = 0.5, minPts = 5)$cluster
  
  data.frame(
    n_neighbors = n,
    ARI = adjustedRandIndex(db1, db2)
  )
}) %>%
  bind_rows()

# Plot stability vs n_neighbors
ggplot(stability_results_4of5, aes(x = n_neighbors, y = ARI)) +
  geom_line() +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  theme_classic() +
  labs(title = "UMAP cluster stability vs n_neighbors",
       x = "n_neighbors", y = "Adjusted Rand Index")
## choose the smallest n_neighbors value where ARI > 0.8
#For all params, it says 10, or 30 (more stable)
#For 4of5 params, it says 15

stability_grid <- expand.grid(
  n_neighbors = neighbor_vals,
  min_dist = min_dist_vals
) %>%
  rowwise() %>%
  mutate(ARI = {
    fit1 <- umap(photophys_scaled, n_neighbors = n_neighbors,
                 min_dist = min_dist, random_state = 42)
    fit2 <- umap(photophys_scaled, n_neighbors = n_neighbors,
                 min_dist = min_dist, random_state = 123)
    
    db1 <- dbscan(fit1$layout, eps = 0.5, minPts = 5)$cluster
    db2 <- dbscan(fit2$layout, eps = 0.5, minPts = 5)$cluster
    
    adjustedRandIndex(db1, db2)
  })

stability_grid_4of5 <- expand.grid(
  n_neighbors = neighbor_vals,
  min_dist = min_dist_vals
) %>%
  rowwise() %>%
  mutate(ARI = {
    fit1 <- umap(photophys_scaled_4of5, n_neighbors = n_neighbors,
                 min_dist = min_dist, random_state = 42)
    fit2 <- umap(photophys_scaled_4of5, n_neighbors = n_neighbors,
                 min_dist = min_dist, random_state = 123)
    
    db1 <- dbscan(fit1$layout, eps = 0.5, minPts = 5)$cluster
    db2 <- dbscan(fit2$layout, eps = 0.5, minPts = 5)$cluster
    
    adjustedRandIndex(db1, db2)
  })

# Visualize as heatmap
ggplot(stability_grid_4of5, aes(x = factor(n_neighbors), 
                                y = factor(min_dist), 
                                fill = ARI)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(ARI, 2)), size = 3) +
  scale_fill_gradient2(low = "white", mid = "lightyellow",
                       high = "darkgreen", midpoint = 0.6) +
  theme_classic() +
  labs(title = "UMAP stability grid search",
       x = "n_neighbors", y = "min_dist")
#This says 0.1,30 for all
# and .01,10 OR .05,50 for _4of5

## tuning eps to umap scale 
# Find optimal eps using k-nearest neighbor distance plot
knn_dist <- kNNdist(umap_fit$layout, k = 5)
plot(sort(knn_dist), type = "l",
     xlab = "Points sorted by distance",
     ylab = "5-NN distance",
     main = "k-NN distance plot for eps selection") ## optimal value is elbow 
# Run DBSCAN with chosen eps
db <- dbscan(umap_fit$layout, 
             eps = 0.5,      # replace with your elbow value
             minPts = 5)     # minimum cluster size

umap_coords <- umap_coords %>%
  mutate(Cluster = as.factor(db$cluster))
# Remember cluster 0 = noise points

# Plot
ggplot(umap_coords, aes(x = UMAP1, y = UMAP2, 
                        color = Cluster, 
                        shape = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic() +
  labs(title = "UMAP with DBSCAN clusters")
```
### final UMAP plotting code  
```{r}
# Final UMAP with optimized parameters -- replace with your optimal parameters!!! 
#This says 0.1,30 for all
# and .01,10 OR .05,50 for _4of5

umap_final <- umap(photophys_scaled_4of5,
                   n_neighbors = 10,
                   min_dist = 0.01,
                   n_components = 2,
                   random_state = 42)  # set seed for reproducibility

umap_coords <- as.data.frame(umap_final$layout) %>%
  setNames(c("UMAP1", "UMAP2")) %>%
  mutate(
    Genotype = allphoto$CoralID,
    Species = allphoto$Species
  )

# Run DBSCAN on final UMAP — rerun kNN plot to tune eps
knn_dist <- kNNdist(umap_final$layout, k = 5)
# Plot the kNN distance curve to visually inspect
data.frame(
  index = 1:length(sort(knn_dist)),
  distance = sort(knn_dist)
) %>%
  ggplot(aes(x = index, y = distance)) +
  geom_line() +
  theme_classic() +
  labs(title = "k-NN distance plot",
       x = "Points sorted by distance",
       y = "5-NN distance")
# Try a range of smaller eps values
eps_vals <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8)

eps_results <- lapply(eps_vals, function(e) {
  db <- dbscan(umap_final$layout, eps = e, minPts = 5)
  data.frame(
    eps = e,
    n_clusters = length(unique(db$cluster[db$cluster != 0])),
    n_noise = sum(db$cluster == 0)
  )
}) %>%
  bind_rows()

print(eps_results)

# Calculate rate of change to find elbow mathematically
sorted_dists <- sort(knn_dist)
rate_of_change <- diff(sorted_dists)

# Elbow is where rate of change jumps most dramatically
elbow_idx <- which.max(rate_of_change)
eps_suggestion <- sorted_dists[elbow_idx]
cat("Suggested eps:", eps_suggestion)

# Run DBSCAN with optimized parameters
db <- dbscan(umap_final$layout, 
             eps = 0.6,
             minPts = 4)

# Check cluster assignments
table(db$cluster)
# cluster 0 = noise points not assigned to any cluster

# Add cluster assignments to umap_coords
umap_coords <- umap_coords %>%
  mutate(Cluster = as.factor(db$cluster),
         SampleID= rownames(umap_coords),
         Site = as.numeric(str_extract(SampleID, "(?<=t)\\d+(?=_)")), 
         Colony =(str_extract(SampleID, "(?<=_).+?(?=\\.)")),
         Species =as.factor(str_to_upper(substr(Colony,1,4))),
         Genotype=paste0(str_to_upper(Species),
                         "_s",formatC(Site,flag=0,width=2),
                         "_c",formatC(as.numeric(substr(Colony,5,6)),flag=0,width=2)))

# Plot
u <- ggplot(umap_coords, aes(x = UMAP1, y = UMAP2,
                             color = Cluster,
                             shape = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic() +
  labs(title = "UMAP with DBSCAN clusters",
       x = "UMAP1", y = "UMAP2")

u

umap_coords=umap_coords %>% left_join(ed,by="Genotype")
umap_coords$SP.C=factor(paste0(umap_coords$Species,".",umap_coords$Cluster))
library(lmerTest)

dt=sort(table(umap_coords$SP.C))
dt=dt[dt<3]
drop_SP.C=which(umap_coords$SP.C%in%c(names(dt)))
if(length(drop_SP.C)==0){MD=umap_coords}else{MD=umap_coords[-drop_SP.C,]}
mod=lmer(ED50~1+(1|Site),data=MD)
modSP=lmer(ED50~Species+(1|Site),data=MD)
modSP_CL=lmer(ED50~Species+Cluster+(1|Site),data=MD)
modSP.C=lmer(ED50~SP.C+(1|Site),data=MD)

anova(mod,modSP,modSP_CL,modSP.C)
anova(modSP,modSP_CL)

#odtest=lm(ED50~Species,data=ed50_test)
smod=summary(modSP);smod
smod=summary(modSP_CL);smod
Spmn=as.data.frame(emmeans(object = modSP_CL,specs = ~ Cluster))
Spmn

```

## Identifying metrics that are significantly different across clusters: random forest importance analysis 
```{r}

# Join cluster assignments to scaled spectral data
rf_data <- photophys_scaled %>%
  mutate(Cluster = as.factor(umap_coords$Cluster))

# Fit random forest
rf_fit <- randomForest(Cluster ~ ., 
                       data = rf_data,
                       ntree = 500,
                       importance = TRUE)

# Extract variable importance
importance_df <- as.data.frame(importance(rf_fit)) %>%
  rownames_to_column("Metrics") %>%
  arrange(desc(MeanDecreaseGini))

# Plot top contributing metrics
importance_df %>%
  head(30) %>%
  ggplot(aes(x = Metrics, y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_classic() +
  labs(title = "Spectral features distinguishing UMAP clusters",
       x = "Photophysiology Metric",
       y = "Mean decrease in Gini impurity")

# high mean decrease in Gini means its consistently important for distinguishing clusters 

### getting predictions of how metrics vary across clusters 
# Step 1 — use random forest to identify top discriminating variables
# reducing from 512 to a manageable set
top_vars <- importance_df %>%
  top_n(30, MeanDecreaseGini) %>%
  pull(Metrics)

# Step 2 — run Kruskal-Wallis only on those top variables 
# this avoids multiple comparisons across all 512 variables 
###### this may not be the best method for yall-- this is what I have been using on my fluorescent protein dataset
kw_results <- photophys_scaled %>%
  as.data.frame() %>%
  mutate(Cluster = as.factor(db$cluster)) %>%
  filter(Cluster != 0) %>%
  select(Cluster, all_of(top_vars)) %>% 
  pivot_longer(-Cluster,
               names_to = "Variable",
               values_to = "Value") %>%
  group_by(Variable) %>%
  kruskal_test(Value ~ Cluster) %>%
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05)

# Step 3 — post-hoc Dunn test for significant variables
posthoc_results <- photophys_scaled %>%
  as.data.frame() %>%
  mutate(Cluster = as.factor(db$cluster)) %>%
  filter(Cluster != 0) %>%
  select(Cluster, all_of(kw_results$Variable)) %>%
  pivot_longer(-Cluster,
               names_to = "Variable",
               values_to = "Value") %>%
  group_by(Variable) %>%
  dunn_test(Value ~ Cluster, p.adjust.method = "BH")

### random forest 

# Plot top 30 variables by importance
v <- importance_df %>%
  top_n(30, MeanDecreaseGini) %>%
  ggplot(aes(x = reorder(Metrics, MeanDecreaseGini),
             y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_classic() +
  labs(title = "Top 30 metrics by random forest importance",
       x = "Metrics",
       y = "Mean decrease in Gini impurity")

#ggsave( v , file = 'Figures/randomforest_importance.png', height = 6, width = 6)
```

# Linear Discriminant Analysis 
For looking at which metrics differ the most across predetermined groups. Would be useful for comparing metrics that vary across sites or species. 
```{r}
dats <- allphoto
phys_lda_data <- dats %>% dplyr::select(where(~!all(is.na(.)))) %>% 
  mutate(Genotype = as.factor(Genotype), 
         Species = as.factor(Species)) %>% 
  mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) %>% na.omit()

# 2. RUN THE LDA MODEL
# Assuming your clean dataframe is called 'phys_lda_data' 
# and your grouping column is called 'Species'

# The ~ . formula tells R: "Predict Species using ALL other columns in the dataset (ignore Genotype)"
lda_phys_model <- lda(Species ~ . - Genotype, data = phys_lda_data)

# Print the model to see the "Coefficients of Linear Discriminants"
# (This tells you which metrics were the most important for separating the groups!)
print(lda_phys_model)

# 3. EXTRACT THE SCORES FOR PLOTTING
# We have to "predict" where each coral falls on the new LD1 axis
lda_predictions <- predict(lda_phys_model)

# Create a new dataframe for ggplot containing the grouping variable and the new LD1 score
phys_lda_plot_data <- data.frame(
  Species = phys_lda_data$Species,
  LD1 = lda_predictions$x[,1] # Grabs the LD1 scores
)

# 4. BUILD THE DENSITY PLOT

l <- ggplot(phys_lda_plot_data, aes(x = LD1, fill = Species)) +
  # geom_density creates those smooth mountain peaks. alpha makes them see-through!
  geom_density(alpha = 0.6) + 
  
  theme_bw() +
  labs(
    title = "LDA: Separating Samples of Different Species",
    # subtitle = "Complete separation of peaks indicates a highly predictive model",
    x = "Linear Discriminant 1 (LD1)",
    y = "Density"
  ) ; print(l)

ggsave(plot = l, file = 'Figures/photophys_LDA.png')
```

