---
title: "Dendrogram American Samoa 2026"
author: "Sophie Wong, Kira Turnham, Brittany Huntington, Courtney Couch, Edo Sena, Kenneth Hoadley"
date: "02/17/2026"
output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
---

# Setup

```{r Setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
rm(list=ls())
```


This script creates a dendrogram of the bio-optical phenotype data from the Fiji 2025 expedition.

### Load Packages
```{r, echo=F}
##Install Packages if Needed
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")

##Load Packages
library(ggplot2) #Required for ggplots
library(dplyr)
library(gplots)
library(gridExtra)
library(car)
library(vegan)
library(reshape)
library(reshape2)
library(broom)
library(pvclust)
library(dendextend)
library(colorspace)
library(circlize)
library(viridis)
library(caret)
library(tidyverse)
library(tidyr)
library(igraph)
library(fs)

```

### Changing file names 
```{r renaming raw files}
# Read the mapping file
sample_metadata <- read.csv(here("Data", "AS26_MAPPmetadata_expt.csv"), check.names = F)
#sample_metadata <- sample_metadata %>% na.omit()
#depth_metadata <- read.csv(file = 'Outputs/AS26_DepthGradientmetadata.csv')
#sample_metadata <- sample_metadata %>% full_join(depth_metadata)
```

# Making dendrogram 

```{r dendro}
photophys <- read.csv("Outputs/AS26_PhenotypingData_expt.csv", row.names = 1)
#photodeep <- read.csv('Outputs/AS26_DepthGradientPhenotypingData.csv', row.names =1)
#photophys <- full_join(photophys, photodeep)

longphoto <- as.data.frame(t(photophys))
colnames(longphoto) <- longphoto[1,] 
longphoto <- as.data.frame(longphoto[-1:-5,]) # removing metrix, stage, col 
longphoto[,2:955] <- lapply(longphoto[,2:955], as.numeric)
longphoto$SampleID <- rownames(longphoto)
longphoto <- longphoto %>% relocate(SampleID)

#remove the duplicate site 7s
longphoto <- longphoto %>%
filter(
    # Look for ICRA or AHYA anywhere in the name, AND 005 at the end
    !(startsWith(SampleID, "t7") & grepl("icra|ahya", SampleID) & endsWith(SampleID, "005r.processed.csv")),
    
    # Look for AGLO or AABR anywhere in the name, AND 002 at the end
    !(startsWith(SampleID, "t7") & grepl("aglo|aabr", SampleID) & endsWith(SampleID, "002r.processed.csv"))
  )

longphoto$SampleID_clean <- longphoto$SampleID %>%
  tolower() %>%
  sub("\\.processed\\.csv$", "", .) %>%
  sub("\\._?\\d{2,3}r$", "", .) %>%   # handles ._02r OR .002r
  trimws()

#check if duplicates
any(duplicated(longphoto$SampleID_clean))


valid_ids <- sample_metadata$UniqueID

longphoto <- longphoto %>%
  filter(SampleID_clean %in% valid_ids)

#longphoto <- longphoto %>%
#  filter(!grepl("^[aoiAOI]", SampleID_clean))

#save this long format data

write.csv(longphoto, "Outputs/longphoto_clean.csv", row.names = FALSE)
```


```{r dendro}
fullyears <- longphoto 


fullyears <- data.frame(fullyears[, colSums(is.na(fullyears)) == 0])

fullyears <- fullyears %>% select(-contains(c('Fo', "Fm"))) # removing Fo and Fm metrics because already included in Quant calculations. Script used to also remove "Con", but since instrument variability were normalized, don't need to remove.
dim(fullyears)
Alls.cntn<-fullyears
Alls.transit<-preProcess(x=Alls.cntn, rangeBounds = c(0,1), method="range")
Alls.trans<-predict(Alls.transit, Alls.cntn)
Alls.preproc<-Alls.trans[,-nearZeroVar(Alls.trans)] # sometimes can throw an error, could be excluded 

dim(Alls.preproc)
Alls.cor<-Alls.preproc 
Alls.cor[,2:ncol(Alls.cor)] <- lapply(Alls.cor[,2:ncol(Alls.cor)], as.numeric)
Alls.cor <- Alls.cor[,-1]
#Alls.cor <- na.omit(Alls.cor) #had to remove this becuae it resulted in 0 for everything
# Remove columns with all NAs or zero variance
bad_cols <- sapply(Alls.cor, function(x) all(is.na(x)) || sd(x, na.rm = TRUE) == 0)
Alls.cor_clean <- Alls.cor[, !bad_cols]
descrCor<-cor(Alls.cor_clean)

```


```{r dendro}
cbb<-pvclust(t(Alls.cor_clean), method.dist="canberra", parallel=TRUE, method.hclust="ward.D", nboot=100)
save(cbb, file = "Outputs/cbb_AS26_expt.RData")
```


```{r dendro}
##### label color coding ####### 
load("Outputs/cbb_AS26_expt_1000bs.RData")
# Create a named color vector (each unique species, i.e. SiteID gets a color)
#site_ids <- unique(guam_mapplist$mapp_no)
site_ids <- unique(sample_metadata$Species)

#site_colors <- setNames(magma(length(site_ids)), site_ids)
my_custom_palette <- c( "chartreuse3", "violetred3", "sienna2", "orchid") 
# This links "Species A" -> "DarkGreen", etc.
site_colors <- setNames(my_custom_palette[1:length(site_ids)], site_ids)

# Get the label order from the dendrogram
dend_labels <- labels(as.dendrogram(cbb))
#dend_labels <- labels(as.dendrogram(cbb))
# Match SiteIDs to dendrogram label order
#label_sites <- trimws(sample_metadata$Species[match(dend_labels, sample_metadata$SampleID)])
# Get the color for each label based on its SiteID
#label_colors <- site_colors[label_sites]
#cb <- color_labels(as.dendrogram(cbb), col = label_colors)


# Convert to dendrogram
d_obj <- as.dendrogram(cbb)
# 3. Use the cleaned labels to match your SiteIDs
dend_labels <- labels(d_obj)
#dend_labels_clean <- sub("_002$", "", dend_labels)
#dend_labels_clean <- sub("\\._(02|05)r$", "", dend_labels_clean)
dend_labels_clean <- dend_labels %>%
    tolower() %>%
  sub("\\.processed\\.csv$", "", .) %>%
  sub("\\._?\\d{2,3}r$", "", .) %>%   # handles ._02r OR .002r
  trimws()

colnames(sample_metadata) <- make.names(colnames(sample_metadata), unique = TRUE)

#sample_metadata <- sample_metadata %>%
  #filter(`CBASS / Sophie` == "CBASS") %>%
 # filter(!is.na(`CBASS / Sophie`)) %>%
#  filter(!is.na(UniqueID)) %>%
#  filter(!grepl("^[aoiAOI]", UniqueID))

sample_metadata$UniqueID <- sample_metadata$UniqueID %>%
  tolower() %>%
  trimws()

label_sites <- sample_metadata$Species[
  match(dend_labels_clean, sample_metadata$UniqueID)
]
label_colors <- site_colors[label_sites]
label_colors[is.na(label_colors)] <- "black"
d_obj <- set(d_obj, "labels_col", label_colors)

#only analyze data here from experiment
#label_sites <- trimws(sample_metadata$Species[match(dend_labels_clean, sample_metadata$UniqueID)])


# Overwrite the actual label text for ease of viewing/reading
labels(d_obj) <- labels(d_obj) %>%
  tolower() %>%
  sub("\\.processed\\.csv$", "", .) %>%   # remove file suffix
  sub("\\.\\d{3}r$", "", .) %>%           # remove .002r / .005r
  trimws()

#make sure there are no duplicates
table(labels(d_obj)) %>% sort(decreasing = TRUE) %>% head()
```


```{r dendro}
#labels_cex(cb)<-0.8
labels_cex(d_obj) <- 0.8
samp<-"Figures/Dendro-Species_expt_circle.pdf"
pdf(file = samp, width = 12, height = 12, bg="transparent")
#circlize_dendrogram(as.dendrogram(cb), labels_track_height = 0.05, dend_track_height = 0.45, labels=TRUE)
circlize_dendrogram(d_obj,labels_track_height = 0.05,dend_track_height = 0.45)
legend("topright", legend = names(site_colors), fill = site_colors, title = "Species")
dev.off()

labels_cex(d_obj) <- 0.3
samp<-"Figures/Dendro-Species_expt_tree.pdf"
pdf(file = samp, width = 10, height = 12)
plot(d_obj, horiz = TRUE,
  main = "Dendrogram - Species",
  ylab = "Height")
legend(
  "topleft",
  legend = names(site_colors),
  fill = site_colors,
  title = "Species",
  cex = 0.8)
dev.off()

```

```{r}
# 2. Match Depths to the dendrogram label order
library(dendextend)
library(viridis)
library(scales)
# We pull the numeric depth for each sample in the order they appear in the tree
depth_vector <- as.numeric(sample_metadata$Colony.Depth..ft.[match(dend_labels, sample_metadata$SampleID)])

# 3. Create a Color Gradient function
# 'magma' or 'viridis' is great for depth; 'reverse = TRUE' makes deep water darker
col_fun <- colorRampPalette(c("cyan", "darkblue")) 

# 4. Map the depths to the colors
# We scale the depths to a range of 0-100 to pick colors from the gradient
scaled_depths <- round(rescale(depth_vector, to = c(1, 100)))
label_colors <- col_fun(100)[scaled_depths]

cb <- color_labels(as.dendrogram(cbb), col = label_colors)

labels_cex(cb)<-0.6
samp<-"Figures/Dendro-Depth_nocon_all.pdf"
pdf(file = samp, width = 12, height = 12, bg="transparent")
circlize_dendrogram(as.dendrogram(cb), labels_track_height = 0.05, dend_track_height = 0.45, labels=TRUE)
# legend("topright", legend = names(scaled_depths), fill = label_colors, title = "Site")
dev.off()

```

# Figures

### Figure Title
Write figure caption.
Note: Adjust figure size if needed.

Circular bar graph 
``` {r standgfp} 
################### 
#orderbleach<-data.frame(fullyears[match(labels(cb), fullyears$SampleID),])
orderbleach<-data.frame(maxwavelength_bycol[match(labels(cb), maxwavelength_bycol$CoralID),])

data3<-orderbleach
dim(data3)
#colnames(data3)
data3$new<-ifelse(data3$standGFP.3 > 1, 1, 0)

empty_bar <- 1
to_add<-matrix(NA, empty_bar, ncol(data3))
colnames(to_add)<-colnames(data3)
data3<-rbind(data3, to_add)
data3<-rbind(data3, to_add)
data3<-rbind(data3, to_add)
data3$id<-seq(1,dim(data3)[1],1)

label_data <- data3
 
number_of_bar <- nrow(label_data)
angle <-  90 - 360 * (label_data$id-0.5)/number_of_bar    
label_data$hjust<-ifelse( angle < -75, 1, 0)
label_data$angle<-ifelse(angle < -75, angle+180, angle)

p <- ggplot(data3, aes(x=as.factor(id), y=-standGFP.3*10, fill=as.factor(new))) +   
  geom_bar(stat="identity") +
  scale_fill_manual("legend", values=c("0"="chocolate3", "1"="chartreuse3")) +
  ylim(500,-100) +
  
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-2,4), "cm")
    	   ) +
  
  coord_polar(start = 0) #+
  #geom_text(data=label_data, aes(x=id, y=-100, label=rownames(data3), hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.25, angle= label_data$angle, inherit.aes = TRUE) 
  # geom_text(data=label_data, aes(x=id, y=-100, label="", hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.25, angle= label_data$angle, inherit.aes = TRUE) + 
  # geom_text(
  #   label=rownames(data3), 
  #   nudge_x = 0.25, nudge_y = 0.25, angle= label_data$angle,
  #   check_overlap = T, size = 1.5
  # )
  p
samp<-paste("Figures/photo/standgfp3_circularbar_gfpdendro", ".pdf", sep="")
pdf(file = samp, width = 10, height = 10, bg="transparent")
p
dev.off()


```


# Significant photophysiology metrics based on bleaching metrics 
## Identifying significantly correlated photophys to bleaching metrics 
```{r}
### just for my own ease of interpretation, I pull out the "meta metrics" which describe more about the relative changes in the photophys traces throughout the light protocol, but these can and should stay in if you want to make the dendrogram about photophysiological differences across colonies/species 
metrics_to_use_nometa <- grep("L|D|Comp|Camp|range|mean|max|min|slope", colnames(lightresponse), invert = T, value = T) 
lightresponse <- photophys %>% select(metrics_to_use_nometa)
metrics_to_use <- metrics_to_use_nometa[-1]

######## 

FinDa <- lightresponse %>% 
  right_join(bleach_metrics)

FinDa <- FinDa %>% dplyr::select(-contains(c('Fm', 'Fo'))) # remove these two metrics because Quant is calculated from them 
FinDa <- FinDa %>% dplyr::filter(!is.na(Quant.1.0 ))

FinDa <- FinDa %>%  column_to_rownames("Genotype")
FinDa <- FinDa %>% dplyr::select(where(~!all(is.na(.))))

   # Get the names of the columns to correlate
all_vars <- FinDa %>% colnames(.)
# Get the names of the non-response metric columns
other_vars <- setdiff(all_vars, metrics_to_use)
# grouping by treatment 

# Create an empty list to store the results
cor_results <- list()

# Loop through each fp_data variable
for (photo_var in metrics_to_use) {
  # For each photo_data variable, loop through all other variables
  for (other_var in other_vars) {
    
    # Calculate the correlation and p-value
    test_result <- cor.test(
      FinDa[[photo_var]], 
      FinDa[[other_var]], 
      method = "spearman", 
      use = "complete.obs"
    )
    
    # Store the results in the list
    cor_results <- append(cor_results, list(data.frame(
      Var1 = photo_var,
      Var2 = other_var,
      Correlation = test_result$estimate,
      PValue = test_result$p.value
    )))
  }
}

# Combine the list of dataframes into a single dataframe
cor_df <- bind_rows(cor_results)

############ removing nonsignificant variables from dataframe to reduce bonferroni correction 
sigsubset <- cor_df %>% 
 dplyr::filter(abs(Correlation) > 0.3) 

sigsubset <- sigsubset %>%
  group_by(Var2) %>%
  mutate(PValue_BY = p.adjust(PValue, method = "BY"), # use BH for photophys analyses and BY for other comparisons 
         PValue_BH = p.adjust(PValue, method = 'BH')) %>% 
  ungroup()

# Now, filter for significant correlations using the new adjusted p-value
significant_correlations <- sigsubset %>%
  dplyr::filter(PValue_BH < 0.05) %>%
  #dplyr::filter(abs(Correlation) > 0.2) %>%
  dplyr::rename(from = Var1, to = Var2, weight = Correlation)
write.csv(significant_correlations, 'Outputs/photophys_bleachingresponse_BHpvaladj.csv', row.names = F)

##### can also use permutation based correlation threshold instead of pvalue 
```

## Bootstrap clustering support across dendrogram branches 
``` {r}
##### pvclust will calculate bootstrap values for you!! just change nboot to >1000 to increase replication 
cbb<-pvclust(t(Alls.cor_clean), method.dist="canberra", parallel=TRUE, method.hclust="ward.D", nboot=1000) 

##### incorporate AU values to find # of significant clusters (AU > 0.95) 
signif_clusters <- pvpick(cbb, alpha = 0.95)$clusters  #30 clusters!
signif_clusters <- pvpick(cbb, alpha = 0.99)$clusters  #17 clusters!

dend_with_p <- as.dendrogram(cbb)

dend_with_p <- color_branches(dend_with_p, k = 4) # use the number of significant clusters from pvpick

plot(dend_with_p)
##### plotting dendrogram ####### 


labels(dend_with_p) <- labels(dend_with_p) %>%
  tolower() %>%
  sub("\\.processed\\.csv$", "", .) %>%   # remove file suffix
  sub("\\.\\d{3}r$", "", .) %>%           # remove .002r / .005r
  trimws()


dend_with_p <- set(dend_with_p, "labels_col", label_colors)

samp<-"Figures/Dendro-by_4k.pdf"
pdf(file = samp, width = 11, height = 11, bg="transparent")
#circlize_dendrogram(dend_with_p, labels_track_height = 0.05, dend_track_height = 0.45, labels=TRUE)
plot(dend_with_p, horiz = TRUE,
  main = "Dendrogram - Species",
  ylab = "Height")
#legend(
 # "topleft",
  #legend = names(Group),
 # fill = site_colors,
  #title = "Group",
  #cex = 0.7)
dev.off() 

samp<-"Figures/Dendro-by_7k_circle.pdf"
pdf(file = samp, width = 12, height = 12, bg="transparent")
circlize_dendrogram(dend_with_p, labels_track_height = 0.05, dend_track_height = 0.45, labels=TRUE)
legend("topright", legend = names(site_colors), fill = site_colors, title = "Species")
dev.off() 
```
```{r}
# make branches use custom cluster colors instead of species-like colors

dend_plot <- dend_with_p

# smaller labels
dend_plot <- dendextend::set(dend_plot, "labels_cex", 0.45)

# custom branch colors for 4 groups
branch_cols <- c(
  "black",
  "darkgoldenrod4",
  "grey45",
  "steelblue4"
  
)


dend_plot <- dendextend::color_branches(
  dend_plot,
  k = 4,
  col = branch_cols
)

# keep labels colored by species
dend_plot <- dendextend::set(dend_plot, "labels_col", label_cols)

pdf("Figures/Dendro-by_4k.pdf", width = 11, height = 11)

plot(
  dend_plot,
  horiz = TRUE,
  main = "Branches = Cluster | Labels = Species",
  ylab = "Height"
)

legend(
  "topleft",
  legend = names(species_cols),
  fill = species_cols,
  title = "Species",
  bty = "n",
  cex = 0.8
)

legend(
  "bottomleft",
  legend = paste("Cluster", 1:4),
  lwd = 2,
  col = branch_cols,
  title = "Tree Groups",
  bty = "n",
  cex = 0.8
)

dev.off()
```

```{r}
library(dendextend)

# --- start from your dendrogram object ---
# Example: dend_with_p already exists

# 1. Make sample labels smaller

dend_plot <- as.dendrogram(dend_with_p)


dend_plot <- dendextend::set(dend_plot, "labels_cex", 0.35)
dend_plot <- dendextend::color_branches(dend_plot, k = 4)

# 3. Color LABELS by species
# assumes labels look like t1_aabr1._02r etc

labs <- labels(dend_plot)

species <- ifelse(grepl("aabr", labs, ignore.case = TRUE), "AABR",
           ifelse(grepl("aglo", labs, ignore.case = TRUE), "AGLO",
           ifelse(grepl("ahya", labs, ignore.case = TRUE), "AHYA",
           ifelse(grepl("icra", labs, ignore.case = TRUE), "ICRA", NA))))

species_cols <- c(
  AABR = "chartreuse3",
  AGLO = "violetred3",
  AHYA = "sienna2",
  ICRA = "orchid"
)

label_cols <- species_cols[species]

# custom branch colors for 4 groups
branch_cols <- c("black", "grey50", "darkred", "steelblue4")
dend_plot <- dendextend::color_branches(
  dend_plot,
  k = 4,
  col = branch_cols
)
dend_plot <- dendextend::set(dend_plot, "labels_cex", 0.45)
dend_plot <- dendextend::set(dend_plot, "labels_col", label_cols)



# 4. Save figure
samp <- "Figures/Dendro-by_4k.pdf"
```


```{r}
pdf("Figures/Dendro-by_4k.pdf", width = 11, height = 11)

plot(
  dend_plot,
  horiz = TRUE,
  main = "Branches = Cluster | Labels = Species"
)

legend(
  "topleft",
  legend = names(species_cols),
  fill = species_cols,
  cex = 0.8,
  bty = "n"
)

dev.off()
```

