```{r}
rm(list = ls())

# ============================================================
# Load data
# ============================================================

library(randomForest)
library(tidyverse)
library(here)
library(pheatmap)
library(rstatix)
library(tidyr)
library(dplyr)

load(here("Data", "Alls.cor_clean.RData"))

load(here("Data",
          "cbb_AS26_expt_1000bs_updated.RData"))

load(here("Data",
          "permanova_matrix_clean.Rdata"))

load(here("Data",
          "annotation_col.Rdata"))


clusters <- cutree(
  cbb$hclust,
  k = 3
)

rownames(Alls.cor_clean)[1:5]

names(clusters)[1:5]

rf_data <- Alls.cor_clean %>%
  mutate(
    Cluster = factor(clusters[rownames(Alls.cor_clean)])
  ) %>%
  select(where(~sum(is.na(.))==0))

rf_data <- Alls.cor_clean %>%
  as.data.frame() %>%
  mutate(
    Cluster = factor(clusters[rownames(.)])
  )

rf_data <- rf_data %>%
  filter(!is.na(Cluster))

rf_predictors <- rf_data %>%
  select(-Cluster) %>%
  scale() %>%
  as.data.frame()


rf_data_scaled <- rf_predictors %>%
  mutate(
    Cluster = rf_data$Cluster
  )

#rf
set.seed(123)

#remove NAs except cluster
predictors <- rf_data_scaled %>%
  select(-Cluster) %>%
  select(where(~ sum(is.na(.)) == 0))

rf_data_scaled_clean <- predictors %>%
  mutate(Cluster = rf_data_scaled$Cluster)
sum(is.na(rf_data_scaled_clean))


rf_fit <- randomForest(
  Cluster ~ .,
  data = rf_data_scaled_clean,
  ntree = 1000,
  importance = TRUE
)


rf_fit

importance_accuracy <- importance(
  rf_fit,
  type = 1
)


importance_accuracy <- importance_accuracy %>%
  as.data.frame() %>%
  rownames_to_column("Metric") %>%
  arrange(desc(MeanDecreaseAccuracy))


head(importance_accuracy,30)

importance_gini <- importance(
  rf_fit,
  type = 2
)


importance_gini <- importance_gini %>%
  as.data.frame() %>%
  rownames_to_column("Metric") %>%
  arrange(desc(MeanDecreaseGini))

top30 <- importance_accuracy %>%
  slice(1:30)

#plot

ggplot(top30,
       aes(x = reorder(Metric, MeanDecreaseAccuracy),
           y = MeanDecreaseAccuracy)) +
  geom_col(fill="steelblue") +
  coord_flip() +
  theme_classic() +
  labs(
    title="Photophysiological metrics driving hclust clusters",
    x="Metric",
    y="Mean decrease in accuracy"
  )

write.csv(
  importance_accuracy,
  here("Data",
       "randomforest_metric_importance.csv"),
  row.names=FALSE
)



# top_vars <- importance_accuracy %>%
#   slice(1:50) %>%
#   pull(Metric)

# top_vars <- top30$Metric

top_vars <- importance_accuracy %>%
  slice(1:100) %>%
  pull(Metric)


kw_results <- rf_data_scaled_clean %>%
  select(Cluster, all_of(top_vars)) %>%
  pivot_longer(
    cols = -Cluster,
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  group_by(Metric) %>%
  kruskal_test(Value ~ Cluster) %>%
  adjust_pvalue(method="BH") %>%
  arrange(p.adj)


kw_results

sig_metrics <- kw_results %>%
  filter(p.adj < 0.05) %>%
  pull(Metric)


dunn_results <- rf_data_scaled_clean %>%
  select(Cluster, all_of(sig_metrics)) %>%
  pivot_longer(
    cols=-Cluster,
    names_to="Metric",
    values_to="Value"
  ) %>%
  group_by(Metric) %>%
  dunn_test(
    Value ~ Cluster,
    p.adjust.method="BH"
  )


dunn_results

rf_fit
```
```{r}
library(randomForest)
library(dplyr)


rf_importance_repeat <- replicate(
  50,
  {
    rf <- randomForest(
      Cluster ~ .,
      data = rf_data_scaled_clean,
      ntree = 1000,
      importance = TRUE
    )
    
    importance(rf, type=1)[,1]
  }
)


importance_stability <- as.data.frame(rf_importance_repeat)

importance_stability$Metric <- rownames(importance_stability)


importance_stability <- importance_stability %>%
  mutate(
    times_selected = rowSums(
      apply(
        select(., -Metric),
        2,
        function(x) x >= quantile(x,0.95)
      )
    )
  ) %>%
  arrange(desc(times_selected))
```



## Identifying metrics that are significantly different across clusters: random forest importance analysis 
```{r}
longphoto <- longphoto %>% relocate(SampleID)
fullyears <- longphoto 
fullyears <- data.frame(fullyears[, colSums(is.na(fullyears)) == 0])

allphoto <- fullyears %>% dplyr::select(-contains('Fo'), -contains('Fm'))

##### dataset needs to have metrics in columns, samples in rows 
rownames(allphoto) <- allphoto$SampleID
photophys_clean <- allphoto %>% 
   dplyr::select(-c(Ramet, CoralID, SampleID)) # remove non-numeric columns

head(photophys_clean) 
names(photophys_clean)[colSums(is.na(photophys_clean)) > 0]
photophys_scaled <- photophys_clean %>%
  scale() %>% # scaling with zscores instead 
  as.data.frame()

# Join cluster assignments to scaled spectral data
rf_data <- photophys_scaled %>%
  mutate(Cluster = as.factor(umap_coords$Cluster))

# Fit random forest
rf_fit <- randomForest(Cluster ~ ., 
                       data = rf_data,
                       ntree = 500,
                       importance = TRUE)

# Extract variable importance
importance_df <- as.data.frame(importance(rf_fit)) %>%
  rownames_to_column("Metrics") %>%
  arrange(desc(MeanDecreaseGini))

# Plot top contributing metrics
importance_df %>%
  head(30) %>%
  ggplot(aes(x = Metrics, y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_classic() +
  labs(title = "Spectral features distinguishing UMAP clusters",
       x = "Photophysiology Metric",
       y = "Mean decrease in Gini impurity")

# high mean decrease in Gini means its consistently important for distinguishing clusters 

### getting predictions of how metrics vary across clusters 
# Step 1 — use random forest to identify top discriminating variables
# reducing from 512 to a manageable set
top_vars <- importance_df %>%
  top_n(30, MeanDecreaseGini) %>%
  pull(Metrics)

# Step 2 — run Kruskal-Wallis only on those top variables 
# this avoids multiple comparisons across all 512 variables 
###### this may not be the best method for yall-- this is what I have been using on my fluorescent protein dataset
kw_results <- photophys_scaled %>%
  as.data.frame() %>%
  mutate(Cluster = as.factor(db$cluster)) %>%
  filter(Cluster != 0) %>%
  select(Cluster, all_of(top_vars)) %>% 
  pivot_longer(-Cluster,
               names_to = "Variable",
               values_to = "Value") %>%
  group_by(Variable) %>%
  kruskal_test(Value ~ Cluster) %>%
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05)

# Step 3 — post-hoc Dunn test for significant variables
posthoc_results <- photophys_scaled %>%
  as.data.frame() %>%
  mutate(Cluster = as.factor(db$cluster)) %>%
  filter(Cluster != 0) %>%
  select(Cluster, all_of(kw_results$Variable)) %>%
  pivot_longer(-Cluster,
               names_to = "Variable",
               values_to = "Value") %>%
  group_by(Variable) %>%
  dunn_test(Value ~ Cluster, p.adjust.method = "BH")

### random forest 

# Plot top 30 variables by importance
v <- importance_df %>%
  top_n(30, MeanDecreaseGini) %>%
  ggplot(aes(x = reorder(Metrics, MeanDecreaseGini),
             y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_classic() +
  labs(title = "Top 30 metrics by random forest importance",
       x = "Metrics",
       y = "Mean decrease in Gini impurity")

#ggsave( v , file = 'Figures/randomforest_importance.png', height = 6, width = 6)
```
 
# Linear Discriminant Analysis 
For looking at which metrics differ the most across predetermined groups. Would be useful for comparing metrics that vary across sites or species. 
```{r}
dats <- allphoto
phys_lda_data <- dats %>% dplyr::select(where(~!all(is.na(.)))) %>% 
  mutate(Genotype = as.factor(Genotype), 
         Species = as.factor(Species)) %>% 
  mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) %>% na.omit()

# 2. RUN THE LDA MODEL
# Assuming your clean dataframe is called 'phys_lda_data' 
# and your grouping column is called 'Species'

# The ~ . formula tells R: "Predict Species using ALL other columns in the dataset (ignore Genotype)"
lda_phys_model <- lda(Species ~ . - Genotype, data = phys_lda_data)

# Print the model to see the "Coefficients of Linear Discriminants"
# (This tells you which metrics were the most important for separating the groups!)
print(lda_phys_model)

# 3. EXTRACT THE SCORES FOR PLOTTING
# We have to "predict" where each coral falls on the new LD1 axis
lda_predictions <- predict(lda_phys_model)

# Create a new dataframe for ggplot containing the grouping variable and the new LD1 score
phys_lda_plot_data <- data.frame(
  Species = phys_lda_data$Species,
  LD1 = lda_predictions$x[,1] # Grabs the LD1 scores
)

# 4. BUILD THE DENSITY PLOT

l <- ggplot(phys_lda_plot_data, aes(x = LD1, fill = Species)) +
  # geom_density creates those smooth mountain peaks. alpha makes them see-through!
  geom_density(alpha = 0.6) + 
  
  theme_bw() +
  labs(
    title = "LDA: Separating Samples of Different Species",
   # subtitle = "Complete separation of peaks indicates a highly predictive model",
    x = "Linear Discriminant 1 (LD1)",
    y = "Density"
  ) ; print(l)

ggsave(plot = l, file = 'Figures/photophys_LDA.png')