# Setup

```{r Setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


This script creates a dendrogram of the bio-optical phenotype data from the compiled file created with ShinyProcessedFile_Compiler.R. 

### Load Packages
```{r, echo=FALSE, include=FALSE}

##Load Packages
library(ggplot2) #Required for ggplots
library(dplyr)
library(gplots)
library(gridExtra)
library(car)
library(vegan)
library(reshape)
library(reshape2)
library(broom)
library(pvclust)
library(dendextend)
library(colorspace)
library(circlize)
library(viridis)
library(caret)
library(tidyverse)
library(tidyr)
library(igraph)
library(fs)
library(alluvial)
library(ggalluvial)
library(umap)     
library(caret)
library(here)
options(scipen=999)
set.seed(123)

```

# Loading Data 
```{r loading data, echo=F}
photophys <- read.csv(here("Data","AS26_PhenotypingData_expt.csv"), row.names = 1)

```

# Trajectory UMAPs 
## Photophys UMAP  

```{r photophys umap setup }
# Load required libraries
library(here)
library(dplyr)
library(caret)
library(vegan) 
library(ggplot2)

# Load data
photophys <- read.csv(here("Data","AS26_PhenotypingData_expt.csv"), row.names = 1)

######### set up photophys dataframe ######## 
longphoto <- as.data.frame(t(photophys))
colnames(longphoto) <- longphoto[1,] 

# --- CRITICAL: Extract your metadata before slicing it away! ---
# We grab SampleID, Site, and spp from longphoto before removing rows 1:5
metadata <- data.frame(
  SampleID = rownames(longphoto)[-1:-5],
  Site = longphoto[-1:-5, "Site"],  # Adjust if the row name in photophys is different (e.g., "site")
  spp = longphoto[-1:-5, "spp"]     # Adjust if the row name in photophys is different (e.g., "Species")
)

longphoto <- as.data.frame(longphoto[-1:-5,]) # removing metrix, stage, col 
longphoto[,2:955] <- lapply(longphoto[,2:955], as.numeric)
longphoto$SampleID <- rownames(longphoto)
longphoto <- longphoto %>% relocate(SampleID)
fullyears <- longphoto 
fullyears <- data.frame(fullyears[, colSums(is.na(fullyears)) == 0])

allphoto <- fullyears %>% dplyr::select(-contains('Fo'), -contains('Fm'))
rownames(allphoto) <- allphoto$SampleID
photophys_clean <- allphoto %>% 
  dplyr::select(-c(SampleID)) # remove non-numeric columns 
photophys_clean <- photophys_clean %>% mutate(across(everything(), as.numeric))
allphoto[,2:ncol(allphoto)] <- allphoto[,2:ncol(allphoto)] %>% mutate(across(everything(), as.numeric))

photo.transit <- preProcess(x = photophys_clean, rangeBounds = c(0, 1), method = "range") # need to scale data 
photo_proc <- predict(photo.transit, photophys_clean)
photo_proc <- photo_proc[,-nearZeroVar(photo_proc)] # remove any columns with low variance across samples 
photo_proc <- photo_proc %>%
  dplyr::select(where(~ sum(is.na(.)) == 0))

# --- NMDS GENERATION ---
# Swapping out the old umap() function for metaMDS
photo_nmdsresult <- metaMDS(photo_proc, 
                            distance = "euclidean", 
                            k = 2, 
                            trymax = 100, 
                            autotransform = FALSE)

# Extract NMDS coordinates
nmds_layout <- as.data.frame(scores(photo_nmdsresult, display = "sites"))
nmds_layout$SampleID <- rownames(nmds_layout)

# Join NMDS coordinates with the metadata extracted above
plot_data <- left_join(nmds_layout, metadata, by = "SampleID")

# --- GGPLOT2 CODE (Color by Site, Shape by spp) ---
ggplot(plot_data, aes(x = NMDS1, y = NMDS2, color = Site, shape = spp)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_bw() +
  labs(
    title = "Photophys Trajectory NMDS",
    subtitle = paste("Stress:", round(photo_nmdsresult$stress, 4)),
    x = "NMDS Dimension 1",
    y = "NMDS Dimension 2",
    color = "Site",
    shape = "Species"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold")
  )

```

# NEW UMAP CODE 
### Load Data and scale 
```{r}
set.seed(123)
longphoto <- as.data.frame(t(photophys))
colnames(longphoto) <- longphoto[1,] 
longphoto <- as.data.frame(longphoto[-1:-5,]) # removing metrix, stage, col 
longphoto[,2:955] <- lapply(longphoto[,2:955], as.numeric)
longphoto$SampleID <- rownames(longphoto)
longphoto <- longphoto %>% relocate(SampleID)
fullyears <- longphoto 
fullyears <- data.frame(fullyears[, colSums(is.na(fullyears)) == 0])

allphoto <- fullyears %>% dplyr::select(-contains('Fo'), -contains('Fm'))

##### dataset needs to have metrics in columns, samples in rows 
rownames(allphoto) <- allphoto$SampleID
photophys_clean <- allphoto %>% 
   dplyr::select(-c(Ramet, CoralID, SampleID)) # remove non-numeric columns

head(photophys_clean) 
names(photophys_clean)[colSums(is.na(photophys_clean)) > 0]
photophys_scaled <- photophys_clean %>%
  scale() %>% # scaling with zscores instead 
  as.data.frame()
```


### evalulating parameters and clustering 
```{r}

# Test range of n_neighbors values
neighbor_vals <- c(5, 10, 15, 20, 30, 50)
min_dist_vals <- c(0.01, 0.05, 0.1, 0.2, 0.5)

umap_list <- lapply(neighbor_vals, function(n) {
  fit <- umap(photophys_scaled, n_neighbors = n, min_dist = 0.1)
  as.data.frame(fit$layout) %>%
    setNames(c("UMAP1", "UMAP2")) %>%
    mutate(n_neighbors = n,
           Genotype = allphoto$CoralID, #add back whatever metadata you want
           Species = allphoto$Species)
})

# Plot all versions for visual comparison
bind_rows(umap_list) %>%
  ggplot(aes(x = UMAP1, y = UMAP2, color = Species)) +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~n_neighbors, labeller = label_both, scales = "free") +
  theme_classic() +
  labs(title = "UMAP stability across n_neighbors values")

# Testing stability of parameter selection: Fit UMAP twice at each neighbor value with different seeds and compare cluster assignments
stability_results <- lapply(neighbor_vals, function(n) {
  
  fit1 <- umap(photophys_scaled, n_neighbors = n, 
               min_dist = 0.1, random_state = 42)
  fit2 <- umap(photophys_scaled, n_neighbors = n, 
               min_dist = 0.1, random_state = 123)
  
  # Cluster both with DBSCAN -- objective clustering method, can change minPts and eps settings (will test eps settings later)
  db1 <- dbscan(fit1$layout, eps = 0.5, minPts = 5)$cluster
  db2 <- dbscan(fit2$layout, eps = 0.5, minPts = 5)$cluster
  
  data.frame(
    n_neighbors = n,
    ARI = adjustedRandIndex(db1, db2)
  )
}) %>%
  bind_rows()

# Plot stability vs n_neighbors
ggplot(stability_results, aes(x = n_neighbors, y = ARI)) +
  geom_line() +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  theme_classic() +
  labs(title = "UMAP cluster stability vs n_neighbors",
       x = "n_neighbors", y = "Adjusted Rand Index")
## choose the smallest n_neighbors value where ARI > 0.8

stability_grid <- expand.grid(
  n_neighbors = neighbor_vals,
  min_dist = min_dist_vals
) %>%
  rowwise() %>%
  mutate(ARI = {
    fit1 <- umap(photophys_scaled, n_neighbors = n_neighbors,
                 min_dist = min_dist, random_state = 42)
    fit2 <- umap(photophys_scaled, n_neighbors = n_neighbors,
                 min_dist = min_dist, random_state = 123)
    
    db1 <- dbscan(fit1$layout, eps = 0.5, minPts = 5)$cluster
    db2 <- dbscan(fit2$layout, eps = 0.5, minPts = 5)$cluster
    
    adjustedRandIndex(db1, db2)
  })

# Visualize as heatmap
ggplot(stability_grid, aes(x = factor(n_neighbors), 
                            y = factor(min_dist), 
                            fill = ARI)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(ARI, 2)), size = 3) +
  scale_fill_gradient2(low = "white", mid = "lightyellow",
                       high = "darkgreen", midpoint = 0.6) +
  theme_classic() +
  labs(title = "UMAP stability grid search",
       x = "n_neighbors", y = "min_dist")

## tuning eps to umap scale 
# Find optimal eps using k-nearest neighbor distance plot
knn_dist <- kNNdist(umap_fit$layout, k = 5)
plot(sort(knn_dist), type = "l",
     xlab = "Points sorted by distance",
     ylab = "5-NN distance",
     main = "k-NN distance plot for eps selection") ## optimal value is elbow 
# Run DBSCAN with chosen eps
db <- dbscan(umap_fit$layout, 
             eps = 0.5,      # replace with your elbow value
             minPts = 5)     # minimum cluster size

umap_coords <- umap_coords %>%
  mutate(Cluster = as.factor(db$cluster))
# Remember cluster 0 = noise points

# Plot
ggplot(umap_coords, aes(x = UMAP1, y = UMAP2, 
                         color = Cluster, 
                         shape = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic() +
  labs(title = "UMAP with DBSCAN clusters")
```
### final UMAP plotting code  
```{r}
# Final UMAP with optimized parameters -- replace with your optimal parameters!!! 
umap_final <- umap(photophys_scaled,
                   n_neighbors = 10,
                   min_dist = 0.01,
                   n_components = 2,
                   random_state = 42)  # set seed for reproducibility

umap_coords <- as.data.frame(umap_final$layout) %>%
  setNames(c("UMAP1", "UMAP2")) %>%
  mutate(
    Genotype = allphoto$CoralID,
    Species = allphoto$Species
  )

# Run DBSCAN on final UMAP — rerun kNN plot to tune eps
knn_dist <- kNNdist(umap_final$layout, k = 5)
# Plot the kNN distance curve to visually inspect
data.frame(
  index = 1:length(sort(knn_dist)),
  distance = sort(knn_dist)
) %>%
  ggplot(aes(x = index, y = distance)) +
  geom_line() +
  theme_classic() +
  labs(title = "k-NN distance plot",
       x = "Points sorted by distance",
       y = "5-NN distance")
# Try a range of smaller eps values
eps_vals <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8)

eps_results <- lapply(eps_vals, function(e) {
  db <- dbscan(umap_final$layout, eps = e, minPts = 5)
  data.frame(
    eps = e,
    n_clusters = length(unique(db$cluster[db$cluster != 0])),
    n_noise = sum(db$cluster == 0)
  )
}) %>%
  bind_rows()

print(eps_results)

# Calculate rate of change to find elbow mathematically
sorted_dists <- sort(knn_dist)
rate_of_change <- diff(sorted_dists)

# Elbow is where rate of change jumps most dramatically
elbow_idx <- which.max(rate_of_change)
eps_suggestion <- sorted_dists[elbow_idx]
cat("Suggested eps:", eps_suggestion)

# Run DBSCAN with optimized parameters
db <- dbscan(umap_final$layout, 
             eps = 0.6,
             minPts = 4)

# Check cluster assignments
table(db$cluster)
# cluster 0 = noise points not assigned to any cluster

# Add cluster assignments to umap_coords
umap_coords <- umap_coords %>%
  mutate(Cluster = as.factor(db$cluster))

# Plot
u <- ggplot(umap_coords, aes(x = UMAP1, y = UMAP2,
                         color = Cluster,
                         shape = Treatment)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic() +
  labs(title = "UMAP with DBSCAN clusters",
       x = "UMAP1", y = "UMAP2")


```

## Identifying metrics that are significantly different across clusters: random forest importance analysis 
```{r}

# Join cluster assignments to scaled spectral data
rf_data <- photophys_scaled %>%
  mutate(Cluster = as.factor(umap_coords$Cluster))

# Fit random forest
rf_fit <- randomForest(Cluster ~ ., 
                       data = rf_data,
                       ntree = 500,
                       importance = TRUE)

# Extract variable importance
importance_df <- as.data.frame(importance(rf_fit)) %>%
  rownames_to_column("Metrics") %>%
  arrange(desc(MeanDecreaseGini))

# Plot top contributing metrics
importance_df %>%
  head(30) %>%
  ggplot(aes(x = Metrics, y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_classic() +
  labs(title = "Spectral features distinguishing UMAP clusters",
       x = "Photophysiology Metric",
       y = "Mean decrease in Gini impurity")

# high mean decrease in Gini means its consistently important for distinguishing clusters 

### getting predictions of how metrics vary across clusters 
# Step 1 — use random forest to identify top discriminating variables
# reducing from 512 to a manageable set
top_vars <- importance_df %>%
  top_n(30, MeanDecreaseGini) %>%
  pull(Metrics)

# Step 2 — run Kruskal-Wallis only on those top variables 
# this avoids multiple comparisons across all 512 variables 
###### this may not be the best method for yall-- this is what I have been using on my fluorescent protein dataset
kw_results <- photophys_scaled %>%
  as.data.frame() %>%
  mutate(Cluster = as.factor(db$cluster)) %>%
  filter(Cluster != 0) %>%
  select(Cluster, all_of(top_vars)) %>% 
  pivot_longer(-Cluster,
               names_to = "Variable",
               values_to = "Value") %>%
  group_by(Variable) %>%
  kruskal_test(Value ~ Cluster) %>%
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05)

# Step 3 — post-hoc Dunn test for significant variables
posthoc_results <- photophys_scaled %>%
  as.data.frame() %>%
  mutate(Cluster = as.factor(db$cluster)) %>%
  filter(Cluster != 0) %>%
  select(Cluster, all_of(kw_results$Variable)) %>%
  pivot_longer(-Cluster,
               names_to = "Variable",
               values_to = "Value") %>%
  group_by(Variable) %>%
  dunn_test(Value ~ Cluster, p.adjust.method = "BH")

### random forest 

# Plot top 30 variables by importance
v <- importance_df %>%
  top_n(30, MeanDecreaseGini) %>%
  ggplot(aes(x = reorder(Metrics, MeanDecreaseGini),
             y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_classic() +
  labs(title = "Top 30 metrics by random forest importance",
       x = "Metrics",
       y = "Mean decrease in Gini impurity")

#ggsave( v , file = 'Figures/randomforest_importance.png', height = 6, width = 6)
```
 
# Linear Discriminant Analysis 
For looking at which metrics differ the most across predetermined groups. Would be useful for comparing metrics that vary across sites or species. 
```{r}
dats <- allphoto
phys_lda_data <- dats %>% dplyr::select(where(~!all(is.na(.)))) %>% 
  mutate(Genotype = as.factor(Genotype), 
         Species = as.factor(Species)) %>% 
  mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) %>% na.omit()

# 2. RUN THE LDA MODEL
# Assuming your clean dataframe is called 'phys_lda_data' 
# and your grouping column is called 'Species'

# The ~ . formula tells R: "Predict Species using ALL other columns in the dataset (ignore Genotype)"
lda_phys_model <- lda(Species ~ . - Genotype, data = phys_lda_data)

# Print the model to see the "Coefficients of Linear Discriminants"
# (This tells you which metrics were the most important for separating the groups!)
print(lda_phys_model)

# 3. EXTRACT THE SCORES FOR PLOTTING
# We have to "predict" where each coral falls on the new LD1 axis
lda_predictions <- predict(lda_phys_model)

# Create a new dataframe for ggplot containing the grouping variable and the new LD1 score
phys_lda_plot_data <- data.frame(
  Species = phys_lda_data$Species,
  LD1 = lda_predictions$x[,1] # Grabs the LD1 scores
)

# 4. BUILD THE DENSITY PLOT

l <- ggplot(phys_lda_plot_data, aes(x = LD1, fill = Species)) +
  # geom_density creates those smooth mountain peaks. alpha makes them see-through!
  geom_density(alpha = 0.6) + 
  
  theme_bw() +
  labs(
    title = "LDA: Separating Samples of Different Species",
   # subtitle = "Complete separation of peaks indicates a highly predictive model",
    x = "Linear Discriminant 1 (LD1)",
    y = "Density"
  ) ; print(l)

ggsave(plot = l, file = 'Figures/photophys_LDA.png')
```

